Variance reduction for MCMC methods via martingale representations

### 2020-02-19 aap/Revision/vr-mcmc-revision-new.tex
update vr-

For multi-indices $\mathbf k,\mathbf k'\in\mathbb N_0^d$
with $\mathbf k'\le\mathbf k$ componentwise
and $\mathbf k'\ne\mathbf k$, $\| k' \| \leq K$, 
we get applying first Lemma~\ref{eq:a_repr},
\begin{equation*}
A_{s,\mathbf{k}}(x)
=\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)^{1/2}
\,
\PE\left[
s^{-1} \sum\nolimits_{r=1}^{s}\{\partial_{1}^{\mathbf{k}'} f(X^x_r)-\PE[\partial_{1}^{\mathbf{k}'}f(X^x_r)]\}\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_1)\right]
\end{equation*}
where $A_{s,\mathbf{k}}$ is defined in \eqref{eq:definition-A-s,k}.
Assume that $\mu$ and $f$ are $K\times d$ times continuously differentiable.
Then, given $\mathbf k\in\mathbb N_0^d$,
by taking $\mathbf k'= \mathbf k'(\mathbf k)
=K(\indiacc{k_1>K}\,\ldots, \indiacc{k_d>K})$, we get
\begin{multline}
\label{eq:sum_abar}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)  
=\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)Q_s(\mathbf{k}',\mathbf{k}-\mathbf{k}')\\
= \left\{ \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}\gamma^{|I|K}\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}}\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!} \right\}
\left\{ \sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K}Q_s(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c}) \right\},
\end{multline}
where for any two multi-indices \(\mathbf{r},\) \(\mathbf{q}\) from \(\mathbb{N}_0^d\)
\begin{equation*}
Q_s(\mathbf{r},\mathbf{q})
=
\left\{\PE\left[\frac{1}{s}\sum_{p=1}^{s}\left\{\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)-\PE\left[\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)\right]\right\}\mathbf{H}_{\mathbf{q}}(Z_{1})\right]\right\}^{2}.
\end{equation*}
In \eqref{eq:sum_abar} the first sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}.$
For any subset $I,$ $\mathbb{N}_{I}^{d}$ stands for a set
of multi-indices $\mathbf{m}_{I}$ with elements $m_{i}=0,$ $i\not\in I,$
and $m_{i}\in\mathbb{N},$  $i\in I.$ Moreover, \(I^c=\{1,\ldots,d\}\setminus I\) and \(\mathbb{N}^d_{0,I^c}\) stands for a set
of multi-indices $\mathbf{m}_{I^c}$ with elements $m_{i}=0,$ $i\in I,$
and $m_{i}\in\mathbb{N}_0,$  $i\not\in I.$ Finally, the multi-index \(\mathbf{K}_I\) is defined as $\mathbf{\mathbf{K}}_{I}=(K1_{\{1\in I\}},\ldots,K1_{\{d\in I\}}).$
Applying the estimate
\begin{eqnarray*}
\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!}\leq (1/2)^{|I| K},
\end{eqnarray*}
we get
\begin{eqnarray*}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)&\leq &
\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma/2)^{|I|K}
\\
&& \times\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}} \sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c})
\\
&\leq &
\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma/2)^{|I|K} \sum_{\mathbf{m}\in\mathbb{N}_0^{d}} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}).
\end{eqnarray*}
Now using the consequence
$\sum_{\mathbf{m}\in \mathbb{N}^d_0} \langle\xi,\mathbf{H}_\mathbf{m}(Z_1)\rangle^2\le\langle\xi,\xi\rangle$
of the Parseval's identity,
we derive
\begin{eqnarray*}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)
&\leq & \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma}{2}\right)^{|I|K}
\PVar\left(s^{-1}\sum\nolimits_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
\right)
\end{eqnarray*}
Next we show that under the conditions  of Theorem~\ref{th:mr}
\begin{eqnarray*}
\PVar\left(s^{-1}\sum\nolimits_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
\right)\leq W_s(x)
\end{eqnarray*}
for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$

 For the sake of simplicity we present the proof only in one-dimensional case. First we need to prove several auxiliary results.
\begin{lem}\label{lem:06062018a1}
Let $(x_p)_{p\in\mathbb N_0}$
and $(\epsilon_p)_{p\in\mathbb N}$
be sequences of nonnegative real numbers
satisfying $x_0=\ol C_0$ and
\begin{align}
0&\le x_p\le\alpha_p x_{p-1}+\gamma \epsilon_p,\quad p\in\mathbb N,
\label{eq:06062018a1}\\
0&\le\epsilon_p\le\ol C_1\prod_{k=1}^p \alpha_k^2,\quad p\in\mathbb N,
\label{eq:06062018a2}
\end{align}
where $\alpha_p,\gamma\in(0,1)$, $p\in\mathbb N$,
and $\ol C_0,\ol C_1$ are some nonnegative constants. Assume
\begin{equation}\label{eq:06062018a3}
\gamma \sum_{r=1}^\infty \prod_{k=1}^r \alpha_k\le\ol C_2
\end{equation}
for some constant $\ol C_2$. Then
$$
x_p\le(\ol C_0+\ol C_1\ol C_2)\prod_{k=1}^p \alpha_k,\quad p\in\mathbb N.
$$
\end{lem}

>>>>

For multi-indices $\mathbf k,\mathbf k'\in\mathbb N_0^d$
with $\mathbf k'\le\mathbf k$ componentwise
and $\mathbf k'\ne\mathbf k$, $\| k' \| \leq K$, 
we get applying first Lemma~\ref{eq:a_repr},
A_{s,\mathbf{k}}(x)
=\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)^{1/2}
\,
\PE\left[
s^{-1} \sum\nolimits_{r=1}^{s}\{\partial_{1}^{\mathbf{k}'} f(X^x_r)-\PE[\partial_{1}^{\mathbf{k}'}f(X^x_r)]\}\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_1)\right]
where $A_{s,\mathbf{k}}$ is defined in \eqref{eq:definition-A-s,k}.
Assume that $\mu$ and $f$ are $K\times d$ times continuously differentiable.
Then, given $\mathbf k\in\mathbb N_0^d$,
by taking $\mathbf k'= \mathbf k'(\mathbf k)
=K(\indiacc{k_1>K}\,\ldots, \indiacc{k_d>K})$, we get
\begin{multline}
\label{eq:sum_abar}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)  
=\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)Q_s(\mathbf{k}',\mathbf{k}-\mathbf{k}')\\
= \left\{ \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}\gamma^{|I|K}\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}}\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!} \right\}
\left\{ \sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K}Q_s(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c}) \right\},
\end{multline}
where for any two multi-indices \(\mathbf{r},\) \(\mathbf{q}\) from \(\mathbb{N}_0^d\)
Q_s(\mathbf{r},\mathbf{q})
=
\left\{\PE\left[\frac{1}{s}\sum_{p=1}^{s}\left\{\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)-\PE\left[\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)\right]\right\}\mathbf{H}_{\mathbf{q}}(Z_{1})\right]\right\}^{2}.
In \eqref{eq:sum_abar} the first sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}.$
For any subset $I,$ $\mathbb{N}_{I}^{d}$ stands for a set
of multi-indices $\mathbf{m}_{I}$ with elements $m_{i}=0,$ $i\not\in I,$
and $m_{i}\in\mathbb{N},$  $i\in I.$ Moreover, \(I^c=\{1,\ldots,d\}\setminus I\) and \(\mathbb{N}^d_{0,I^c}\) stands for a set
of multi-indices $\mathbf{m}_{I^c}$ with elements $m_{i}=0,$ $i\in I,$
and $m_{i}\in\mathbb{N}_0,$  $i\not\in I.$ Finally, the multi-index \(\mathbf{K}_I\) is defined as $\mathbf{\mathbf{K}}_{I}=(K1_{\{1\in I\}},\ldots,K1_{\{d\in I\}}).$
Applying the estimate
\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!}\leq (1/2)^{|I| K},
we get
\label{eq:sumA}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)
&\leq 
\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma/2)^{|I|K}
& \times\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}} \sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c})
&\leq 
\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma/2)^{|I|K} \sum_{\mathbf{m}\in\mathbb{N}_0^{d}} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}).
The Parseval identity implies that for any function $\varphi: \rset^d \to \rset$ satisfying $\PE[\varphi^2(Z_1)] < \infty$,
\sum_{\mathbf{m}\in \nset^d_0} \{\PE[\varphi(Z_1) \mathbf{H}_\mathbf{m}(Z_1)] \}^2 \le \PE[\{\varphi(Z_1)\}^2]
Using this identity in \eqref{eq:sumA} implies 
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)
&\leq & \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\PVar\left(s^{-1}\sum\nolimits_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
\right)
Next we show that under the conditions  of Theorem~\ref{th:mr}
\PVar\left(s^{-1}\sum\nolimits_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)

To keep the notational burden at a reasonable level, we present the proof only in one-dimensional case. 
Multidimensional extension is straightforward but requires involved notations. 
First we need to prove several auxiliary results.
\begin{lem}\label{lem:06062018a1}
Let $(x_p)_{p\in\mathbb N_0}$
and $(\epsilon_p)_{p\in\mathbb N}$
be sequences of nonnegative real numbers
satisfying $x_0=\ol C_0$ and
\begin{align}
0&\le x_p\le\alpha_p x_{p-1}+\gamma \epsilon_p,\quad p\in\mathbb N,
\label{eq:06062018a1}\\
0&\le\epsilon_p\le\ol C_1\prod_{k=1}^p \alpha_k^2,\quad p\in\mathbb N,
\label{eq:06062018a2}
\end{align}
where $\alpha_p,\gamma\in(0,1)$, $p\in\mathbb N$,
and $\ol C_0,\ol C_1$ are some nonnegative constants. Assume
\begin{equation}\label{eq:06062018a3}
\gamma \sum_{r=1}^\infty \prod_{k=1}^r \alpha_k\le\ol C_2
\end{equation}
for some constant $\ol C_2$. Then
$$
x_p\le(\ol C_0+\ol C_1\ol C_2)\prod_{k=1}^p \alpha_k,\quad p\in\mathbb N.
$$
\end{lem}
